---
title: "Betting Shops Analysis ReadMe"
format: gfm
always_allow_html: true
---

# Overview
This blog looks at where betting shops are located around the UK to calculate the number of bookmakers per million people by local authority. Comparing this data to the Index of Multiple Deprivation shows how more deprived areas see a higher number of bookies per person.

`scripts/0-create-data.R` 

# Data on bookmakers
The data on the locations of bookmakers comes from Open Street Map. Using the `osmdata` r package, I collected the locations of each bookmaker which Open Street Map has data for.
```{r}
#| eval: false
library(osmdata)
library(sf)
# Get all points for england and wales
bookmakers <- opq("england") |> # includes data for wales
  add_osm_feature(key = "shop", value = "bookmaker") |>
  osmdata_sf()

# Get all points for scotland
bookmakers_scotland <- opq("scotland") |>
  add_osm_feature("shop", "bookmaker") |>
  osmdata_sf()
```

Once this data was collected, I realised there was a problem of massively overestimating the number of bookies. This was caused by the point data representing the 4 corners of the shop building. So instead of having one point per bookmaker, some shops had 4 or more. This is solved by finding all of these points and replacing them with on point in the center of the building. The data from england, wales and scotland is then combined into one dataset.

```{r}
#| eval: false
# function to remove duplicate points for the same bookmaker
remove_dups <- function(x) {
  x$osm_points |>
    st_join(x$osm_polygons |> mutate(building_poly = "yes"), st_intersects) |>
    filter(is.na(building_poly)) |>
    bind_rows(st_centroid(x$osm_polygons))
}

# combine data for Scotland, England and Wales
bookies <- remove_dups(bookmakers_scotland) |>
  bind_rows(remove_dups(bookmakers)) |>
  distinct() |>
  select(name, addr.postcode) |>
  st_transform(27700)
```

For this data to be meaningful for analysis, we need to know which area each of these points is in. I do this by sending the latitude and longitude of each point to the postcodes.io api, which returns information on the local authority, lower super output area, region and country.

```{r}
#| eval: false
# get area data from postcodes.io
get_area_data <- function(x) {
  lons <- unlist(purrr::map(x$geometry, 1))
  lats <- unlist(purrr::map(x$geometry, 2))

  lsoas <- vector(length = length(lons))
  lsoa_names <- vector(length = length(lons))
  lads <- vector(length = length(lons))
  lad_names <- vector(length = length(lons))
  country <- vector(length = length(lons))
  region <- vector(length = length(lons))

  pb <- txtProgressBar(style = 3)
  pb_add <- seq(1/length(lons), 1, 1/length(lons)) # amount to add each loop

  for (i in 1:length(lons)) {
    api_return <- tryCatch(
      jsonlite::fromJSON(glue::glue("https://api.postcodes.io/postcodes?lon={lons[i]}&lat={lats[i]}&limit=1"))$result,
      error = function(e) return(NULL)
    )
    if (is.null(api_return)) {
      lsoas[i] <- NA
      lsoa_names[i] <- NA
      lads[i] <- NA
      lad_names[i] <- NA
      country[i] <- NA
      region[i] <- NA
    } else {
      lsoas[i] <- api_return$codes$lsoa
      lsoa_names[i] <- api_return$lsoa
      lads[i] <- api_return$codes$admin_district
      lad_names[i] <- api_return$admin_district
      country[i] <- api_return$country
      region[i] <- api_return$region
    }
    setTxtProgressBar(pb, pb_add[i])
  }

  out <- x |>
    mutate(
      lsoa = lsoas,
      lsoa_name = lsoa_names,
      area_code = lads,
      area_name = lad_names,
      country = country,
      region = region
    )


  return(out)
}

# Fetch the data and filter out any points which aren't in a lower super output area
bookies <- bookies |>
  get_area_data() |>
  filter(!is.na(lsoa))
```

This leaves us with a point coordinate for each bookmaker in the UK along with variables for what area it is located.

# Merging with data on deprivation
For data on deprivation I used the English Index of Deprivation from the Ministry of Housing Communities and Local Government (now the Department for Leveling Up Housing and Communities). They provide a dataset of the average deprivation score by local authority. I also use population estimates from the ONS to calculate the number of bookies per person
```{r}
#| eval: false

library(dplyr)
library(data.table)

# Get data for deprivation
deprivation_la <- readxl::read_xlsx("betting-shops/data/deprivation-lad.xlsx",
                                 sheet = 2) |>
  select(area_code = `Local Authority District code (2019)`,
         deprivation = `IMD - Average score`,
         imd_rank = `IMD - Rank of average score`) |>
  as.data.table()

uk_pop <- readxl::read_xls("betting-shops/data/uk-population-estimates-mid-2020.xls", sheet = 7, range = "A8:D426") |>
  select(area_code = Code, population = `All ages`) |>
  as.data.table()
```

Using the `data.table` package I merge these datasets together and calculate the number of bookies per million people by local authority.

```{r}
#| eval: false

la_n_bookies <- bookies[, .(n_bookies = .N), by = .(region, area_name, area_code)][
  as.data.table(uk_pop), on = "area_code", nomatch = 0
][
  deprivation_la, on = "area_code"
]
la_n_bookies[, bookies_per_million := round(n_bookies / population * 1e6, 3)]
la_n_bookies[, region_name := case_when(region %in% c("North East", "North West", "Yorkshire and The Humber") ~ "North",
                                        region %in% c("East Midlands", "West Midlands", "East of England") ~ "Midlands and East",
                                        region %in% c("South West", "South East") ~ "South",
                                        TRUE ~ region)]
```

This data is available in `public/betting-shops-deprivation.csv`. It is subject to the [Open Street Map license](https://www.openstreetmap.org/copyright).
